Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-10-26 writeup/review-responses.tex

\begin{point}{Figures S5 and S6:} 
    Include a panel which gives a legend/image to explain the colors being used for the bottom PCA such that the reader can know what it will look like when one of the PCAs in that row is mirroring geography.
\end{point}

\reply{
    Done.
}

\begin{point}{Senior editor comments:}
    It would be helpful to expand the discussion of variation in relationships
    along the genome - as written, the discussion is strongly focused on the
    "classical" Drosophila-centred linked selection literature.  Specifically,
    there has been a lot of discussion in the phylogenetics world about how to
    estimate species trees from the set of locus-specific genealogies (where the
    "consensus" tree can be quite wrong); and there are methods that use the
    distribution of genealogical relationships to estimate population structure
    (Konrad Lohse, Asger Hobolth etc).  It would be helpful to make these broader
    links.  
\end{point}

\reply{
    Good suggestion; we've added (hopefully helpful) additional discussion.
    \revref and \revreffull{1}{2}
}

@@

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    Li \& Ralph in their manuscript show how local PCA can be used to visualise genomic regions that differ from each other in the mean relatedness among samples. While a similar approach has be previously used to detect inversions in human genome (Ma \& Amos 2012), they have applied their method to also other than human datasets and invoke other interpretations to the observed patterns than inversions.

    I appreciate the quality of writing (readability, clarity, organisation) and also the time spent amending the manuscript after the first review. The figures were clear and demonstrated the points they should. The method was very well explained and the code is understandable to use and easily accessible. The manuscript was fun to read.
\end{quote}

Thanks!  We put a lot of time into the work and the writing,
and that's really nice to hear.

\begin{quote}
    I have not previously reviewed the manuscript and but I think that most of the comments of previous reviewers were well reacted upon. However, the resulting manuscript is to a new reader a bit unclear in terms of the application and usability of the method and ability to distinguish between different scenarios leading to observed patterns and I think further simulations of mechanisms of interest should be performed with the aim to either demonstrate what the limitations of the method are for potential users or to further explore biological processes hinted at.
\end{quote}

A similar point was raised by another reviewer in the previous round.
It might be reasonable, as \citet{ma2012investigation} did,
to use these results to develop an inference method for a particular signal.
We have rather chosen to follow the lead of genomic PCA,
which, despite having connections to demographic quantities,
is most widely-used as a purely descriptive tool,
for communicating population genetics resuls.
We've tried to make this a bit more clear in the introduction. \llname{ll:descriptive}

\begin{point}{}
    Is the method meant to facilitate discovery of new inversions? If so, what number of samples are necessary? Of what coverage (what level of missingness is allowed)? What size of inversions is possible to detect? What are ranges of the distance values between windows that could point to inversions (any significance test would be recommended)? How does the method compare to other methods that detect inversions (esp. Ma \& Amos 2012)? If it is not meant to discover new inversions and it should only confirm known existing inversions, what is the success rate (with varying sample size, missingness)?
\end{point}

\reply{
    These are excellent questions for a method designed to detect inversions.
    However, we do not present such a method,
    we rather present a visualization method in which, empirically, inversions often leave a large (but not quantified) signal.
    We are not sure if it would be productive to turn this into a quantitative method, either:
    as we have now clarified in the paper \llname{ll:inversion_caveat},
    distinguishing inversions from regions of low recombination
    likely requires other sources of information.
}

\begin{point}{}
    The authors argue that linked selection is the likely cause for the patterns they observed in Drosophila dataset since the recombination rate correlates with the position of the first MDS coordinate (from line number 321). At the same time, the authors mention that the ability to identify the windows could be influenced by recombination rate in the case of linked selection and also in the presence of recombination hotspots (section 3.1 Validation). As such, I do not agree with the authors, that it is enough to demonstrate that linked selection could cause these patterns but to use this in interpretation, they should be able to distinguish this explanation from the other possible causes in simulations first and then on real data (e.g., using recombination maps for Drosophila to simulate exactly the same regions could be of advantage).
\end{point}

\reply{
    We don't share the reviewer's concern about variation in recombination rate:
    a region of low recombination could be dominated by a single tree,
    and therefore provide a strongly divergent PCA signal,
    there is no reason this would provide a \emph{consistent} signal
    across a large fraction of a chromosome,
    unless that chromosomal region had much lower recombination rate than we see here.
    More generally, we heartily agree that we would love to have available a test
    that could prove to what degree these (or any) varying patterns of relatedness along genomes
    are due to linked selection, demographic noise, or some other force.
    However, this is one of the major unsolved question in the field,
    and is subject to substantial debate 
    (see for instance \citet{jensen2014unfounded,harris2018unfounded}
    and subsequent rejoinders).
    We've added more to the discussion on this. \revref
}

@@

\begin{point}{}
    The authors refer to supplementary Figures sometimes as "Figure S6" and sometimes as "Supplementary Figure S6". In certain journals, these two could denote different figures so I would select one way only.
\end{point}

\reply{
    Good point; that's fixed.
}

\begin{point}{}
    I do not understand while when plotting the positions, the authors opted to show 2 graphs (for each MDS coordinate) when identifying outliers along the genome. While it is connected to the MDS plot they also show, I would rather try to express this in 1 value because I think that it would increase their ability to detect the outliers (e.g. Figure S6 - the 2 top left figures showing results of simulations with constant recombination rate when compared to MDS coordinate 1 and 2 - if the information from the two graphs was combined in 1 measure, I would expect the pattern to be more visible).
\end{point}

\reply{
    That's a good suggestion -- perhaps, to test for (or, visualize) outlierness,
    one could plot the deviation from the MDS centroid.
    We chose to plot each MDS coordinate separately
    so as to better see how the structure of the MDS plot lays out along the genome.
    For instance, in Figure \ref{fig:mds_allchr} on chromosome arm 3L,
    it is interesting that the green arm of the MDS plot localizes to the endpoints of the inversion,
    while the purple arm localizes to near the centromere.
    If only a measure of outlierness were plotted,
    it would be more difficult to discern patterns like this,
    and therefore develop hypotheses explaining the observed patterns.
}

@@

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    The authors have put in a considerable effort to address the concerns of the reviewers. The simulation studies are now much better and the conclusions were weakened somewhat in light of them. The method definitely appears to be useful. I think that it could be an important part of the literature that moves us forward in the empirical understanding of how the genome evolves.
\end{quote}

\begin{point}{}
I only have one major comment, which is unfortunate because it could have been avoided if the simulation were implemented slightly differently. The authors didn't address the intention of my concern about the switching of PCs. To be very explicit: when you perform a PC analysis, there is some variability about the order that the "true" directions of the data comes out in terms of the magnitude of that PC (i.e. the order). This model uses K PCs for each small segment of genome, and each small segment is compared using a method called MDS. It is convincing that order doesn't affect comparisons of the segments *provided that all of the components are present*. However, if the magnitude of the noise is high, then it is quite plausible that the top K components in a segment of genome are not the top K for the whole genome. The simulation is useless for reassuring on this because it sets up a situation with 2 important PCs which are always included, and the claim that "the simulation also verifies insensitivity to ordering of top PCs" is not really true. The real data analyses are more reassuring, given the relatively similar results between K=2 and K=5, but I think the concern remains valid.

I would like to see evidence that the windows that are found by this method are not associated with "which" PCs happen to be included in the top K. To do that, you could for example calculate the top $M>K$ pcs, check the correlation with the overall PCs, and then plot "presence" of a PC (measured in correlation in the top K) against the MDS values. It would alternatively be possible to make an argument that it will always be fine if you use a large enough number of PCs, but this is argued to have severe computational problems (an argument I don't know that I buy). It would be enough to do this for retaining *one* PC in the simulation that has 2 *equally strong* PCs, but I have a suspicion that this case goes quite wrong and is anyway unrealistic. Better to keep more, e.g. 5 when there is non-trivial structure in the truth, and see that the 4th or 5th swapping with the 6th or 7th doesn't affect much. To be clear: the authors shouldn't necessarily follow this route, but they should address the underlying concern.
\end{point}

\reply{
    Apologies for not realizing the precise nature of your original concern.
    Yes, we agree that if strongly different PCs switched order 
    right at the boundary of which PCs were included,
    then that could easily show up as (spurious) structure in our MDS plots.
    We have added discussion of this point. \revref
    However, for our empirical data we think we have strong evidence that the differences between windows 
    are not being driven by this effect:
    as the reviewer also mentions, the effect is expected to dissappear when using a larger number of PCs,
    which does not happen in our applications.
    Furthermore, and speculatively,
    our method of choosing a window size looks for a big enough window that the PCs within a window 
    are relatively stable,
    and so might help avoid sensitive PC switching.
}

### 2018-10-26 writeup/local_pca_paper.tex
responses

\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
        \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
        \renewcommand\includegraphics{\relax}
        \renewcommand{\includegraphics}[2][]{\relax}
        \renewcommand{\figurename}{Fig.}
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    \newcommand{\Figure}{Fig }
    \newcommand{\Figures}{Fig }
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} 
    % \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
    \newcommand{\Figure}{{Figure }}
    \newcommand{\Figures}{{Figures }}
    % \renewcommand{\llabel}[1]{}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

@@

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
(e.g., \citet{turner2005genomic,ellegren2012genomic}).
However, a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
which in retrospect is unsurprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
Our aim is not to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome;
correlation of this variation with known genomic features
can then be used to uncover its source. % \revpoint{2}{1}

@@

\paragraph{Human:}
We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
of 3,965 samples in total: 346 African-Americans, 73 Asians, 3,187 Europeans and 359 Indian Asians.
Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
which are each derived from whole genome sequencing.
We excluded the sex chromosomes and the mitochondria.
PCA plots for each chromosome, separately, are shown in \Figure \ref{sfig:pca_human_allchr}.
% We use the allele that has highest frequency in the samples as the reference allele for each position.

@@

\paragraph{\textit{Medicago truncatula}:}
Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
which has 263 samples from 24 countries,
primarily distributed around the Mediterranean basin.
Each of the 8 chromosomes has 3--5 million SNPs;
PCA plots for these are shown in \Figure \ref{sfig:pca_medicago_allchr}.
We did not use the mitochondria or chloroplasts.

@@

Simulations with linked selection produced mixed results
(Figure \ref{sfig:slim_sims2}).
The method strongly identified the regions under spatially varying linked selection.
It also identified the regions (although less unambiguously)
with constant selection and stepwise varying recombination rate,
but did not clearly identify them with constant recombination rate.
This difference is likely because recombination rates are overall lower in the first case,
leading to a stronger effect of linked selection.
These tests are not meant to be comprehensive survey of linked selection,
but only to demonstrate that linked selection can produce signals similar to what we see in real data.

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \Figure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), % \revpoint{1}{3}
and are only weakly correlated with variation in missingness 
(see \Figure \ref{sfig:dpgp_missingness}). % \revpoint{2}{4}

@@

We also applied the method on all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal (\Figure \ref{sfig:mds1_along_allchr_human}). 
Further investigation with a denser set of SNPs,
allowing a finer genomic resolution,
may yield other patterns.

@@

@@

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). % \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information.

@@

Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
and how many MDS coordinates to use when describing the distance matrix between windows,
but in our exploration, changing these has not produced dramatically different results.
These are all part of more general techniques in dimension reduction and high-dimensional data visualization;
we encourage the user to experiment.

@@

We first created ``genotypes'' in this way with fifty individuals from each of the three populations;
running our method on a genome with 99 windows of 400 loci each
produced the first plot in Figure \ref{sfig:test_missing}.
These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
and so the ordering of the top two PCs is arbitrary.
If our method was sensitive to PC ordering,
then half the windows in each region that have one ordering would cluster with each other,
separate from the other half. \llabel{ll:pc_switching_test}

We then marked each genotype in the first half of the chromosome
as missing, independently, with probability $1/2$ and ran our method again,
producing the second plot of Figure \ref{sfig:test_missing}. \llabel{ll:test_missing}
If our method was influenced by missing data,
we would expect the first half of the chromosome to separate from the second
in the MDS plot.

>>>>

\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
        \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
        \renewcommand\includegraphics{\relax}
        \renewcommand{\includegraphics}[2][]{\relax}
        \renewcommand{\figurename}{Fig.}
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    \newcommand{\Figure}{Fig }
    \newcommand{\Figures}{Fig }
    \newcommand{\SFigure}{Sup Fig }
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} 
    % \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
    \newcommand{\Figure}{{Figure }}
    \newcommand{\Figures}{{Figures }}
    \newcommand{\SFigure}{{Supplementary Figure }}
    % \renewcommand{\llabel}[1]{}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

@@

It is common to describe variation along the genome of simple statistics such as $F_{ST}$
and to interpret the results in terms of the action of selection 
(e.g., \citet{turner2005genomic,ellegren2012genomic}).
However, a given pattern (e.g., valleys of $F_{ST}$)
can be caused by more than one biological process \citep{cruickshank2014reanalysis,burri2015linked},
which in retrospect is unsurprising given that we are using a single statistic
to describe a complex process.
It is also common to use methods such as PCA to visualize large-scale patterns in 
mean genome-wide relatedness.
In this paper we show if and how patterns of mean relatedness vary systematically along the genome, 
in a way particularly suited to large samples from geographically distributed populations.
Geographic population structure sets the stage by establishing ``background'' patterns of relatedness;
our method then describes how this structure is affected by selection and other factors.
The method is descriptive:
it does not aim to identify outlier loci,
but rather to describe larger-scale variation shared by many parts of the genome,
and give clues about its source. \llabel{ll:descriptive}
% If such variation is found, correlation of this variation with known genomic features
% can then be used to uncover its source. % \revpoint{2}{1}

@@

\paragraph{Human:}
We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
of 3,965 samples in total: 346 African-Americans, 73 Asians, 3,187 Europeans and 359 Indian Asians.
Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
which are each derived from whole genome sequencing.
We excluded the sex chromosomes and the mitochondria.
PCA plots for each chromosome, separately, are shown in \SFigure \ref{sfig:pca_human_allchr}.
% We use the allele that has highest frequency in the samples as the reference allele for each position.

@@

\paragraph{\textit{Medicago truncatula}:}
Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
which has 263 samples from 24 countries,
primarily distributed around the Mediterranean basin.
Each of the 8 chromosomes has 3--5 million SNPs;
PCA plots for these are shown in \SFigure \ref{sfig:pca_medicago_allchr}.
We did not use the mitochondria or chloroplasts.

@@

Simulations with linked selection produced mixed results
(\SFigure \ref{sfig:slim_sims2}).
The method strongly identified the regions under spatially varying linked selection
(also our simulation with strongest positive selection). \revpoint{2}{2}
It also identified the regions (although less unambiguously)
with constant selection and stepwise varying recombination rate,
but did not clearly identify them with constant recombination rate.
This difference is likely because recombination rates are overall lower in the first case,
leading to a stronger effect of linked selection.
These tests are not meant to be comprehensive survey of linked selection,
but only to demonstrate that linked selection can produce signals similar to what we see in real data.

@@

We applied the method to windows of average length 9 Kbp, across chromosome arms 2L, 2R, 3L, 3R and X separately.
The first column of \Figure \ref{fig:mds_allchr} is a multidimensional scaling (MDS) visualization 
of the matrix of dissimilarities between genomic windows:
in other words, genomic windows that are closer to each other in the MDS plot show more similar patterns of relatedness.
For each chromosome arm, the MDS visualization roughly resembles a triangle,
sometimes with additional points.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme manifestations of population structure 
typified by windows found in the ``corners'' of the figure,
and that other windows' patterns of relatedness may be a mixture of those extremes. 
The next two columns of \Figure \ref{fig:mds_allchr} respectively depict the two MDS coordinates of each window,
plotted against the window's position along the genome,
to show how the plot of the first column is laid out along the genome.
The patterns did not depend on the number of PCs used
(see \SFigure \ref{sfig:dmel_k5} for the same plot with $k=5$ PCs), % \revpoint{1}{3}
and are only weakly correlated with variation in missingness 
(see \SFigure \ref{sfig:dpgp_missingness}). % \revpoint{2}{4}

@@

We also applied the method on all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal (\SFigure \ref{sfig:mds1_along_allchr_human}). 
Further investigation with a denser set of SNPs,
allowing a finer genomic resolution,
may yield other patterns.

@@

\paragraph{Genealogical noise?} \revpoint{AE}{6}
The field of phylogenetics has long had to deal with the fact that 
there can be a great number of different local phylogenies along the genome,
even between species
\citep{avise1983mitochondrial,pamilo1988relationships,hobolth2007genomic}.
Loosely, we are studying the within-species patterns that might lead to such
incomplete lineage sorting in the future.
The neutral distribution of variation in these patterns 
has been used to infer demographic history,
both between species \citep{slatkin2006concordance}
and within species \citep{beeravolu2018blockwise}.
If these distinct phylogenies are merely a result of neutral stochasticity,
there is not expected to be a correlation between local phylogeny
and other genomic features.
However, in some cases the local assortment of ancestral diversity 
and subsequent introgression between sister taxa
shows clear signs of selection, 
as for instance in the wild tomato clade \citep{pease2016phylogenomics}.

@@

There has been substantial debate over the relative impacts of different forms of selection
(e.g., \citet{charlesworth1997effects,charlesworth2012effects,pease2013accurate,hedrick2013adaptive,burri2015linked,corbettdetig2015natural,harris2016genetic,martin2016natural,phung2016determining,stankowski2018tempo}). % \revpoint{2}{16}
These have been difficult to disentangle in part because for the most part
theory makes predictions which are only strictly valid in randomly mating (i.e., unstructured) populations,
and it is unclear to what extent the spatial structure observed in most real populations will affect these predictions.
It may be possible to design more powerful statistics that make stronger use of spatial information,
or even using the variation in relatedness that we observe here,
similar to the method of \citet{beeravolu2018blockwise}. \revpoint{1}{2}

@@

Finally, our software allows different choices for how many PCs to use in approximating structure of each window ($k$ in equation \ref{eqn:est_cov}),
and how many MDS coordinates to use when describing the distance matrix between windows,
but in our exploration, changing these has not produced dramatically different results.
However, this choice could easily be important:
for instance, if the $k^\text{th}$ and $(k+1)^\text{st}$ PCs 
are sufficiently different but have similar eigenvalues,
then small amounts of noise could cause these to switch,
leading to spuriously inferred differences between windows in which one or the other
was included in the top $k$ PCs.
This does not seem to be a problem in our applications,
as changing the number of PCs did not affect the qualitative results. \revpoint{2}{1}
These choices are all part of more general techniques in dimension reduction and high-dimensional data visualization;
we encourage the user to experiment.

@@

We first created ``genotypes'' in this way with fifty individuals from each of the three populations;
running our method on a genome with 99 windows of 400 loci each
produced the first plot in \SFigure \ref{sfig:test_missing}.
These matrices are chosen so that the top two eigenvalues $\Sigma$ are the same (both 50.1),
and so the ordering of the top two PCs is arbitrary.
If our method was sensitive to PC ordering,
then half the windows in each region that have one ordering would cluster with each other,
separate from the other half. \llabel{ll:pc_switching_test}

We then marked each genotype in the first half of the chromosome
as missing, independently, with probability $1/2$ and ran our method again,
producing the second plot of \SFigure \ref{sfig:test_missing}. \llabel{ll:test_missing}
If our method was influenced by missing data,
we would expect the first half of the chromosome to separate from the second
in the MDS plot.
